---
title: Analysis of the knowledge about pain in students of Health Sciences
format:
  html:
    theme:
      light: flatly
      dark: darkly
    self-contained: true
    table-of-contents: true
execute: 
  warning: false
  message: false
---

```{r}
#| echo: false
myred = "#F8766D"
myblue = "#00BFC4"
```

## Objetives

- To evaluate whether there are statistically significant differences in knowledge about pain among students of the Physiotherapy, Medicine, Nursing and Pharmacy degrees.
- To evaluate whether there is statistically significant difference in knowledge about pain between students of the first and fourth year of the degrees.
- To evaluate whether there is an interaction between the course and the degree in knowledge about pain.

## Methodology

Cross-sectional observational study. The following questionnaires have been applied to the participating students:

- RNPQ ([Reliability and Pain Knowledge Questionnaire](https://psycnet.apa.org/doiLanding?doi=10.1037%2Ft77106-000)). Questionnaire with 12 questions with three possible answers (True, False, Don't know). 1 point is scored for each correct answer and 0 points for each incorrect answer or "Don't know". The total score ranges from 0 to 12 points, with higher scores indicating greater knowledge about pain.

- HC-PAIRS ([Health care providers' attitudes and beliefs about functional impairments and chronic back pain](https://pubmed.ncbi.nlm.nih.gov/8788576/)). Questionnaire with 15 questions with 7 possible answers on a liker scale (from Totally disagree to Totally agree). It is scored from 1 to 7 points. The total score ranges from 15 to 105, with lower scores indicating greater knowledge about pain.


```{r}
# Packages loading.
library(tidyverse)
library(knitr)
library(broom)
library(vtable)
library(tidymodels)
library(ggpubr)
library(car)
library(rstatix)
library(emmeans)
library(patchwork)
```

```{r}
# Data loading.
df <- read_csv("datos/datos-2024-11-25.csv") |>
    # Select the variables of interest.
    select(Degree, Course, Gender, Age, RNPQ, HC_PAIRS, starts_with("X"), starts_with("Y")) |> 
    # Convert the categorical variables to factors.
    mutate(Degree = factor(Degree), Course = factor(Course), Gender = factor(Gender)) |> 
    # Create a new variable with the combination of the degree and the course.
    mutate(Degree_Course = interaction(Degree, Course))

# Show 10 random rows of the data frame.
sample_n(df, 10) |> 
    kable()
```

## Sample

### Variables 

```{r}
# Show the structure of the data frame.
vt(df)
```

### Sample size

The number of students who participated in the study was `r nrow(df)`.

Sample size according to the degree.

```{r}
df |> 
  count(Degree) |> 
  kable()
```

Sample size according to the course.

```{r}
df |> 
  count(Course) |> 
  kable()
```

Sample size according to the degree and course.

```{r}
table(df$Degree, df$Course) |> 
  kable()
```

Samples size according to the gender.

```{r}
df |> 
  count(Gender) |> 
  kable()
```
## Descriptive analysis

### Descriptive statistics

```{r}
df  |> 
    summarize(`Age mean` = mean(Age), `Age sd` = sd(Age)) |>
    kable()
```
#### According to the degree

```{r}
library(vtable)
st(df, vars = c("RNPQ", "HC_PAIRS"), group = "Degree", summ = c("mean(x)",   "sd(x)", "min(x)", "pctile(x)[25]", "median(x)", "pctile(x)[75]", "max(x)"), title = "Resumen descriptivo",  group.long = TRUE,  summ.names = c("Mean", "StdDev", "Min", "Q1", "Median", "Q3", "Max"), digits = 3)
```

#### According to the course

```{r}
library(vtable)
st(df, vars = c("RNPQ", "HC_PAIRS"), group = "Course", summ = c("mean(x)",   "sd(x)", "min(x)", "pctile(x)[25]", "median(x)", "pctile(x)[75]", "max(x)"), title = "Resumen descriptivo",  group.long = TRUE,  summ.names = c("Mean", "StdDev", "Min", "Q1", "Median", "Q3", "Max"), digits = 3)
```

#### According to the degree and course

```{r}
library(vtable)
st(df, vars = c("RNPQ", "HC_PAIRS"), group = "Degree_Course", summ = c("mean(x)",   "sd(x)", "min(x)", "pctile(x)[25]", "median(x)", "pctile(x)[75]", "max(x)"), title = "Resumen descriptivo",  group.long = TRUE,  summ.names = c("Mean", "StdDev", "Min", "Q1", "Median", "Q3", "Max"), digits = 3)
```

### Boxplot

```{r}
df |>
    ggplot(aes(x = Degree, y = RNPQ, fill = Degree)) +
    geom_boxplot() +
    labs(title = "Distribution of the RNPQ score by degree",
         x = "Degree",
         y = "Score") +
    theme_minimal()    
```

```{r}
df |>
    ggplot(aes(x = Course, y = RNPQ, fill = Course)) +
    geom_boxplot() +
    labs(title = "Distribution of the RNPQ score by course",
         x = "Course",
         y = "Score") +
    theme_minimal()    
```


```{r}
df |>
    ggplot(aes(x = Degree, y = RNPQ, fill = Degree)) +
    geom_boxplot() +
    labs(title = "Distribution of the RNPQ score by degree and course",
         x = "Degree",
         y = "Score") +
    facet_grid(. ~ Course) +
    theme_minimal()    
```

```{r}
df |>
    ggplot(aes(x = Degree, y = HC_PAIRS, fill = Degree)) +
    geom_boxplot() +
    labs(title = "Distribution of the HC-PAIRS score by degree",
         x = "Degree",
         y = "Score") +
    theme_minimal()    
```

```{r}
df |>
    ggplot(aes(x = Course, y = HC_PAIRS, fill = Course)) +
    geom_boxplot() +
    labs(title = "Distribution of the HC-PAIRS score by course",
         x = "Course",
         y = "Score") +
    theme_minimal()    
```

```{r}
df |>
    ggplot(aes(x = Degree, y = HC_PAIRS, fill = Degree)) +
    geom_boxplot() +
    labs(title = "Distribution of the HC-PAIRS score by degree and course",
         x = "Degree",
         y = "Score") +
    facet_grid(. ~ Course) +
    theme_minimal()    
```
### Barplot

```{r}
df |> 
    ggplot(aes(x = RNPQ, fill = Course)) +
    geom_bar(aes(y = ..count../sum(..count..))) +
    labs(title = "Distribution of the RNPQ score by course",
         x = "RNPQ Score",
         y = "Frequency") +
    facet_grid(Course ~ .) +
    theme_minimal()
```

```{r}
df |> 
    ggplot(aes(x = RNPQ, fill = Degree)) +
    geom_bar(aes(y = ..count../sum(..count..))) +
    labs(title = "Distribution of the RNPQ score by degree",
         x = "RNPQ Score",
         y = "Frequency") +
    facet_grid(Degree ~ .) +
    theme_minimal()
```

### Histogram

```{r}
df |> 
    ggplot(aes(x = HC_PAIRS, fill = Course)) +
    geom_histogram(aes(y = ..count../sum(..count..)), color = "white") +
    labs(title = "Distribution of the HC-PAIRS score by course",
         x = "HC-PAIRS score",
         y = "Frequency") +
    facet_grid(Course ~ .) +
    theme_minimal()
```

```{r}
df |> 
    ggplot(aes(x = HC_PAIRS, fill = Degree)) +
    geom_histogram(aes(y = ..count../sum(..count..)), color = "white") +
    labs(title = "Distribution of the HC-PAIRS score by degree",
         x = "HC-PAIRS score",
         y = "Frequency") +
    facet_grid(Degree ~ .) +
    theme_minimal()
```

### Means plot

```{r}
ggline(df, x = "Course", y = "RNPQ", color = "Degree", add = c("mean_ci"), position = position_dodge(0.2), xlab = "Course", ylab = "RNPQ score", main = "Confidence intervals for the means of RNPQ score by degree and course", legend = "top")
```

:::{.callout-caution}
It looks that the pain knowledge according to RNPQ score decreases in the last year of Nursing degree, that is very weird.
:::

```{r}
ggline(df, x = "Course", y = "HC_PAIRS", color = "Degree", add = c("mean_ci"), position = position_dodge(0.2), xlab = "Course", ylab = "HC-PAIRS score", main = "Confidence intervals for the means of HC-PAIRS score by degree and course", legend = "top")
```

:::{.callout-caution}
It looks that the pain knowledge according to HC_PAIRS decreases in the last year of Pharmacy degree, that is very weird.
:::

## Comparison of RNPQ scores

As the dependent variable si quantitative (discrete) and the dependent variables are qualitative, we perform a two-factors ANOVA test for comparing the means of RNPQ score for the groups based on degree and course.

### ANOVA

```{r}
anova <- aov(RNPQ ~ Degree * Course, data = df)
# Use the type III sum of squares as the groups are unbalanced.
Anova(anova, type = "III") |>
    tidy() |>
    mutate(p.value = format(p.value, scientific = TRUE, digits = 2)) |>
    kable()
```

The ANOVA test shows both, the degree and the course, have a statistically significant effect on the scores of the RNPQ questionnaire, and also the interaction between the degree and the course has a statistically significant effect (p-value < 0.01).

#### ANOVA assumptions

Now we study the residuals of the model to check the ANOVA assumptions are meet.

```{r}
# Add residuals and fitted scores to the data frame.
df <- df |> mutate(
    predictions = fitted(anova),
    residuals = residuals(anova)
)
```

First we check normality of residuals.

```{r}
par(mfrow = c(1, 2))  
hist(df$residuals)
qqnorm(df$residuals, main = "Q-Q Plot of Residuals")
qqline(df$residuals, col = "red", lwd = 2)
```

According to the charts there is no significant departure from normality.

After we check homocedasticity.

```{r}
leveneTest(RNPQ ~ Degree * Course, data = df) |> 
    tidy() |> 
    kable()
```

The variances of the scores of the RNPQ questionnaire are homogeneous in all the degree-course groups (p-value > 0.05).

### Pairwise comparison of means

As the interaction of degree and course is significant, we estimate the mean for each degree-course group.

```{r}
marginal_means <- emmeans(anova, ~ Degree * Course)
marginal_means |> 
    tidy(conf.int = TRUE) |> 
    select(-statistic, -p.value) |> 
    arrange(desc(estimate)) |> 
    kable()
```

```{r}
marginal_means |> 
    as_tibble() |> 
    mutate(DegreeCourse = paste(Degree, Course, sep = "-")) |> 
    mutate(DegreeCourse = reorder(DegreeCourse, emmean)) |> 
    ggplot(aes(x = emmean, y = DegreeCourse)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    labs(title = "Estimated means of RNPQ score",
       x = "Estimated Mean",
       y = "Degree - Course") +
    theme_minimal()
```

Now we compare the means of all the groups by pairs.

```{r}
marginal_means |>
    pairs(infer = T) |>
    as_tibble() |> 
    arrange(p.value) |> 
    kable()
```

As many of this comparisons make no sense, we perform a pairwise comparison of degrees by course.

```{r}
marginal_means|> 
    pairs(infer = T, by = "Course") |>   
    as_tibble() |> 
    arrange(Course, p.value) |> 
    kable()
```

```{r}
pwpp(marginal_means, by = "Course") +
    labs(title = "P-values of pairwise comparisons of the RNPQ means by course")
```

```{r}
plot(marginal_means, by = "Course", comparisons = TRUE) +
    labs(title = "95% confidence intervals for the estimated means\nof RNPQ scores and pairwise comparisons", x = "Estimated means") +
    theme_minimal()
```

```{r}
marginal_means|> 
    pairs(infer = T, by = "Course") |>
    as_tibble() |> 
    filter(Course == "first") |> 
    mutate(contrast = reorder(contrast, estimate)) |> 
    ggplot(aes(x = estimate, y = contrast)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "95% confidence intervals for the difference between the RNPQ means in the last course",
       y = "Degrees") +
    theme_minimal()
```

In the first course there are significant differences between nursing and pharmacy,  pharmacy and physioterapy, medicine and nursing and medicine and pharmacy. 

```{r}
marginal_means|> 
    pairs(infer = T, by = "Course") |>
    as_tibble() |> 
    filter(Course == "last") |> 
    mutate(contrast = reorder(contrast, estimate)) |> 
    ggplot(aes(x = estimate, y = contrast)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "95% confidence intervals for the difference between the RNPQ means in the last course",
       x = "Estimated Mean difference",
       y = "Degrees") +
    theme_minimal()
```

In the last course, there are significant differences between nursing and physiotherapy, medicine and nursing, pharmacy and physiotherapy and medicine and nursing.

```{r}
marginal_means|> 
    eff_size(sigma = sigma(anova), edf = 708, by = "Course") |> 
    as_tibble() |>
    arrange(Course, desc(abs(effect.size))) |>
    kable()
```

The biggest effect size detected was between Physiotherapy and Nursing, followed by Medicine and Nursing, Physiotherapy and Pharmacy and Medicine and Pharmacy, all of them with an effect size d>0.5. 

In the same way, we compare the courses for each degree.

```{r}
marginal_means |> 
    pairs(infer = T, by = "Degree") |> 
    as_tibble() |> 
    arrange(p.value) |> 
    kable()
```

```{r}
pwpp(marginal_means, by = "Degree") +
    labs(title = "P-values of comparisons of the RNPQ means\nof first and last courses by dregee")
```

```{r}
plot(marginal_means, by = "Degree", comparisons = TRUE) +
    labs(title = "95% confidence intervals for the estimated means\nof RNPQ scores and pairwise comparisons", x = "Estimated means") +
    theme_minimal()
```

```{r}
marginal_means|> 
    pairs(infer = T, by = "Degree") |>
    as_tibble() |> 
    mutate(Degree = reorder(Degree, estimate)) |> 
    ggplot(aes(x = estimate, y = Degree)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "Difference between the first and last courses means by degrees",
       x = "Estimated Mean difference",
       y = "Degrees") +
    theme_minimal()
```

There are significant differences between the first and the last courses in medicine, pharmacy and physiotherapy, but not in nursing.

```{r}
marginal_means|> 
    eff_size(sigma = sigma(anova), edf = 708, by = "Degree") |> 
    as_tibble() |>
    arrange(desc(abs(effect.size))) |>
    kable()
```

## Comparison of HC-PAIRS scores

### ANOVA 

```{r}
anova <- aov(HC_PAIRS ~ Degree * Course, data = df)

anova |>
    tidy() |>
    kable()
```

The ANOVA test shows both, the degree and the course, have a statistically significant effect on the scores of the HC-PAIRS questionnaire, and also the interaction between the degree and the course has a statistically significant effect (p-value < 0.01).

#### ANOVA assumptions

Now we study the residuals of the model to check the ANOVA assumptions are meet.

```{r}
# Add residuals and fitted scores to the data frame.
df <- df |> mutate(
    predictions = fitted(anova),
    residuals = residuals(anova)
)
```

First we check normality of residuals.

```{r}
par(mfrow = c(1, 2))  
hist(df$residuals)
qqnorm(df$residuals, main = "Q-Q Plot of Residuals")
qqline(df$residuals, col = "red", lwd = 2)
```

According to the charts there is no significant departure from normality.

After we check homocedasticity.

```{r}
leveneTest(HC_PAIRS ~ Degree * Course, data = df) |> 
    tidy() |> 
    kable()
```

The variances of the scores of the HC_PAIRS questionnaire are homogeneous in all the degree-course groups (p-value > 0.05).

```{r}
par(mfrow = c(1,1))
plot(anova, which=1)
```

The scatter plot of residuals vs predicted scores shows homogeneity of variances.

### Pairwise comparison of means

As the interaction of degree and course is significant, we estimate the mean for each degree-course group.

```{r}
marginal_means <- emmeans(anova, specs =  ~ Degree:Course, weights = "cells")
marginal_means |> 
    tidy(conf.int = TRUE) |> 
    select(-statistic, -p.value) |> 
    arrange(desc(estimate)) |> 
    kable()
```

```{r}
marginal_means |> 
    as_tibble() |> 
    mutate(DegreeCourse = paste(Degree, Course, sep = "-")) |> 
    mutate(DegreeCourse = reorder(DegreeCourse, emmean)) |> 
    ggplot(aes(x = emmean, y = DegreeCourse)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    labs(title = "Estimated means of HC_PAIRS score",
       x = "Estimated Mean",
       y = "Degree - Course") +
    theme_minimal()
```

Now we compare the means of all the groups by pairs.

```{r}
contrast(marginal_means, method = "pairwise") |>
    as_tibble() |> 
    arrange(p.value) |> 
    kable()
```

As many of this comparisons make no sense, we perform a pairwise comparison of degrees by course.

```{r}
emmeans(anova, ~ Degree | Course, weights = "cells") |> 
    pairs() |> 
    as_tibble() |> 
    arrange(Course, p.value) |> 
    kable()
```

```{r}
pwpp(marginal_means, by = "Course") +
    labs(title = "P-values of pairwise comparisons of the HC-PAIRS means by course")
```

```{r}
plot(marginal_means, by = "Course", comparisons = TRUE) +
    labs(title = "95% confidence intervals for the estimated means\nof HC-PAIRS scores and pairwise comparisons", x = "Estimated means") +
    theme_minimal()
```

```{r}
emmeans(anova, ~ Degree | Course) |> 
    pairs() |> 
    confint() |> 
    as_tibble() |> 
    filter(Course == "first") |> 
    mutate(contrast = reorder(contrast, estimate)) |> 
    ggplot(aes(x = estimate, y = contrast)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "Difference of HC_PAIRS means in the first course",
       x = "Estimated Mean difference",
       y = "Degrees") +
    theme_minimal()
```

In the first course there are significant differences between pharmacy and physiotherapy and between nursing and physioterapy. 

```{r}
emmeans(anova, ~ Degree | Course) |> 
    pairs() |> 
    confint() |> 
    as_tibble() |> 
    filter(Course == "last") |> 
    mutate(contrast = reorder(contrast, estimate)) |> 
    ggplot(aes(x = estimate, y = contrast)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "Difference of HC_PAIRS means in the last course",
       x = "Estimated Mean difference",
       y = "Degrees") +
    theme_minimal()
```

In the last course, there are significant differences between pharmacy and physiotherapy, nursing and physiotherapy, medicine and pharmacy, medicine and physiotherapy and nursing and pharmacy.

```{r}
marginal_means|> 
    eff_size(sigma = sigma(anova), edf = 708, by = "Course") |> 
    as_tibble() |>
    arrange(Course, desc(abs(effect.size))) |>
    kable()
```

In the same way, we compare the courses for each degree.

<!-- ```{r}
contrast(marginal_means, method = "pairwise", by = "Degree", adjust = "tukey")
``` -->

```{r}
emmeans(anova, ~ Course | Degree, weights = "cells") |> 
    pairs() |> 
    as_tibble() |> 
    arrange(p.value) |> 
    kable()
```

```{r}
pwpp(marginal_means, by = "Degree") +
    labs(title = "P-values of comparisons of the HC-PAIRS means\nof first and last courses by dregee")
```

```{r}
plot(marginal_means, by = "Degree", comparisons = TRUE) +
    labs(title = "95% confidence intervals for the estimated means\nof HC-PAIRS scores and pairwise comparisons", x = "Estimated means") +
    theme_minimal()
```

```{r}
emmeans(anova, ~ Course | Degree) |> 
    pairs() |> 
    confint() |> 
    as_tibble() |> 
    mutate(Degree = reorder(Degree, estimate)) |> 
    ggplot(aes(x = estimate, y = Degree)) +
    geom_errorbar(aes(xmin = lower.CL, xmax = upper.CL), size = 1, width = 0.2, color = myblue) +
    geom_point(size = 3, color = myred) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(title = "Difference between the first and last courses means by degrees",
       x = "Estimated Mean difference",
       y = "Degrees") +
    theme_minimal()
```

There are significant differences between the first and the last courses in pysiotherapy and medicien, but not in nursing and pharmacy.


```{r}
marginal_means|> 
    eff_size(sigma = sigma(anova), edf = 708, by = "Degree") |> 
    as_tibble() |>
    arrange(desc(abs(effect.size))) |>
    kable()
```