---
title: Study about pain knowledge and attitudes among students in health sciences degrees
date: today
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
# Packages loading.
library(tidyverse)
library(knitr)
library(broom)
library(vtable)
library(tidymodels)
library(ggpubr)
library(car)
library(rstatix)
library(emmeans)
library(patchwork)
```

```{r}
# Data loading.
df <- read_csv("datos/datos-2024-11-25.csv") |>
    # Select the variables of interest.
    select(Degree, Course, Gender, RNPQ, HC_PAIRS, starts_with("X"), starts_with("Y")) |> 
    # Convert the categorical variables to factors.
    mutate(Degree = factor(Degree), Course = factor(Course), Gender = factor(Gender)) |> 
    # Create a new variable with the combination of the degree and the course.
    mutate(Degree_Course = interaction(Degree, Course))
```

## Materials and Methods

### Study design and participants

We planned a cross-sectional study including students from first and last courses of Health Sciences degrees (Medicine, Physiotherapy, Nursing and Pharmacy) at several universities of Spain. 

The participants were informed about the nature of the study, agreed to participate, and provided and
informed consent according to the data protection law currently in force. They were asked to complete two self-administered questionnaires, the Spanish version of the Reliability and Pain Knowledge Questionnaire (RNPQ) [referencia], and the Health Care Providers’ Pain and Impairment Relationship Scale (HC-PAIRS) [referencia].

Inclusion and exclusion criteria.

### Statistical analysis

Data processing and analysis were performed with programmes R (version 3.5), RKWard
(version 0.8.0) and rk.Teaching package (version 1.4) [1].

A descriptive summary of the main variables in the study was performed by groups (degrees and courses), including the mean and standard deviation.

The comparison among groups was performed using a two-factors ANOVA test where the factors were the degree and the course. The post-hoc analysis was performed using the Tukey's test. The effect size was calculated using the eta-squared statistic. The assumptions for the ANOVA test were checked: normality of residuals was proven with histogram and Q-Q plot of residuals, and the homogeneity of variances was tested with Levene’s test. We considered a significance level of $\alpha$=0.05.

The required sample size was calculated with the `pwr` package in R, considering a power of 0.8 and a significance level of 0.05. The effect size was calculated using the Cohen’s d statistic assuming a medium effect size. For this requirements we got a minimum sample size of 60 participants per group.

<!-- The comparison between the first and last courses was performed using the Student’s t-test for independent samples. The effect size was calculated using Cohen’s d. -->

## Results

We got a total of 716 participants from the four degrees. The distribution of participants by degree and course is shown in Table 1.

```{r}
#| label: tab:sample-size
#| tbl-cap: Sample size by degree and course
table(df$Degree, df$Course) |> 
  kable()
```

```{r}
table(df$Gender)
```

```{r}
st(df, vars = c("RNPQ", "HC_PAIRS"), group = "Degree_Course", summ = c("mean(x)",   "sd(x)", "min(x)", "pctile(x)[25]", "median(x)", "pctile(x)[75]", "max(x)"), title = "Mean and standard deviation of the RNPQ and HC_PAIRS scores by degree and course")
```

```{r}
anova <- aov(RNPQ ~ Degree * Course, data = df)
# Use the type III sum of squares as the groups are unbalanced.
Anova(anova, type = "III")  |> 
    tidy() |>
    mutate(p.value = format(p.value, scientific = TRUE, digits = 2)) |>
    kable()
```


## References

[1] Sánchez-Alberca, A. Bringing R to non-expert users with the package RKTeaching. Boletín de Estadística e
Investigación Operativa (BEIO) 2015, 31, 170–188.